---
title: "Silhouette Package"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Silhouette Package}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r check-packages, echo=FALSE, message=FALSE, warning=FALSE}
required_packages <- c("proxy", "ppclust", "blockcluster", "cluster", "factoextra", "ggplot2")

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  message("❌ The following required packages are not installed:\n")
  message(paste0("- ", missing_packages, collapse = "\n"), "\n")
  message("📦 To install them, run the following in R:\n")
  message("```r")
  message(sprintf("install.packages(c(%s))", paste(sprintf('\"%s\"', missing_packages), collapse = ", ")))
  message("```")
  
  knitr::knit_exit()
}
```

```{r setup,echo=FALSE, include=FALSE}
library(Silhouette)
library(proxy)
library(ppclust)
library(cluster)
library(factoextra)
library(blockcluster)
library(ggplot2)
library(drclust)
set.seed(123)
```

# Introduction

The Silhouette package provides a comprehensive and extensible framework for computing and visualizing silhouette widths to assess clustering quality in both crisp (hard) and soft (fuzzy/probabilistic) clustering settings. Silhouette width, originally introduced by @rousseeuw1987silhouettes, quantifies how similar an observation is to its assigned cluster relative to the closest alternative cluster. Scores range from -1 (indicative of poor clustering) to 1 (excellent separation).

*Note:* This package does not use the classical @rousseeuw1987silhouettes calculation directly. Instead, it generalizes and extends silhouette methodology as follows:

- Implements the Simplified Silhouette method [@vanderlaan2003pam], with options for medoid or pac (Probability of Alternative Cluster, @raymaekers2022silhouettes) approaches.
- Provides soft clustering silhouettes based on membership probabilities [@campello2006fuzzy; @bhatkapu2024density], including density-based diagnostics like certainty and density-based silhouettes.
- Includes density-based silhouette (dbSilhouette) computation, which leverages log-ratios of posterior probabilities for soft clustering evaluation [@menardi2011density].
- Supports calculation of crisp, fuzzy, and median silhouette widths, allowing flexible averaging methods to suit different clustering needs.
- Supports multi-way clustering evaluation via extSilhouette() [@schepers2008multimode], enabling silhouette analysis for biclustering or higher-order tensor clustering.
Offers customizable and informative visualization with plotSilhouette(), including grayscale options and detailed cluster legends. The package also integrates with clustering results from popular R packages such as cluster (silhouette, pam, clara, fanny) and factoextra (eclust, hcut).
- Includes utility functions for creating and validating Silhouette objects directly from components.

This vignette demonstrates the essential features of the package using the well-known iris dataset. It showcases both standard (crisp) and fuzzy silhouette calculations, advanced plotting capabilities, and extended silhouette metrics for multi-way clustering scenarios.

## Available Functions

**`Silhouette()`**: Calculates silhouette widths for both crisp and fuzzy clustering, using user-supplied proximity matrices.

**`softSilhouette()`**: Computes silhouette widths tailored to soft clustering by interpreting membership probabilities as proximities.

**`dbSilhouette()`**: Computes density-based silhouette widths for soft clustering, based on log-ratios of posterior probabilities.

**`cerSilhouette()`**: Computes certainty silhouette widths for soft clustering, using the maximum posterior probabilities as silhouette values.

**`getSilhouette()`**: Constructs a Silhouette class object directly from user-provided components (e.g., cluster assignments, neighbor clusters, silhouette widths).

**`is.Silhouette()`**: Tests whether an object is of class "Silhouette", with optional strict structural validation.

**`plot() / plotSilhouette()`**: Visualizes silhouette widths as sorted bar plots, offering grayscale and flexible legend options for clarity.

**`summary()`**: Produces concise summaries of average silhouette widths and cluster sizes for objects of class Silhouette.

**`extSilhouette()`**: Derives extended silhouette widths for multi-way clustering problems, such as biclustering or tensor clustering.

# Use Cases

## 1. Simplified Silhouette Calculation

### a. When the Proximity Matrix is Unknown but Centers of Clusters Are Known

This example demonstrates how to compute silhouette widths for a clustering result when you have the proximity (distance) matrix between observations and cluster centres unknown.
The workflow uses the classic `iris` dataset and k-means clustering.

**Steps:**

-   **Clustering**: Perform k-means clustering on `iris[, -5]` with 3 clusters.

```{r kmeans}
data(iris)
km <- kmeans(iris[, -5], centers = 3)
```

*Note*: The `kmeans` output (`km`) does not include a proximity matrix.
Therefore, distances between observations and cluster centroids must be computed separately.

-   **Compute the Proximity Matrix**:\
    Create a matrix of distances between each observation and cluster centroid using `proxy::dist()`.

```{r crisp-silhouette1, fig.width=7, fig.height=4, fig.alt = "fig1.1"}
library(proxy)
dist_matrix <- proxy::dist(iris[, -5], km$centers)
sil <- Silhouette(dist_matrix)
head(sil)
summary(sil)
plot(sil)
```

-   **Customize Calculation**:\
    To use the Probability of Alternative Cluster (PAC) method (which is more penalised variation of medoid method) and return a sorted output:

```{r crisp-silhouette2, fig.width=7, fig.height=4, fig.alt = "fig1.2"}
sil_pac <- Silhouette(dist_matrix, method = "pac", sort = TRUE)
head(sil_pac)
summary(sil_pac)
plot(sil_pac)
```

-   **Accessing Silhouette Summaries**:\
    The `Silhouette` function prints overall and cluster-wise silhouette indices to the R console if `print.summary = TRUE`, but these values are not directly stored in the returned object. To extract them programmatically, use the `summary()` function:

```{r crisp-silhouette3}
s <- summary(sil_pac,print.summary = TRUE)
# summary table
s$sil.sum
# cluster wise silhouette widths
s$clus.avg.widths
# Overall average silhouette width
s$avg.width
```

### b. When the Proximity Matrix Is Known

This section describes how to compute silhouette widths when the **proximity matrix**—representing distances between observations and cluster centers—is readily available as part of the clustering model output.
The example makes use of fuzzy c-means clustering via the `ppclust` package and the classic `iris` dataset.

**Steps:**

-   **Step 1: Perform Fuzzy C-Means Clustering**\
    Apply fuzzy c-means clustering on `iris[, -5]` to create three clusters.

```{r fm}
library(ppclust)
data(iris)
fm <- ppclust::fcm(x = iris[, -5], centers = 3)
```

-   **Step 2: Compute Silhouette Widths Using the Proximity Matrix**\
    The output object `fm` contains a distance matrix `fm$d` representing proximities between each observation and each cluster center, which can be directly fed to the `Silhouette()` function.

```{r crisp-silhouette4, fig.width=7, fig.height=4, fig.alt = "fig1.2"}
sil_fm <- Silhouette(fm$d)
plot(sil_fm)
```


- **Alternative: Directly Use the Clustering Function with `clust_fun`**\
    To streamline the workflow, you can let the `Silhouette()` function internally handle both clustering and silhouette calculation by supplying the name of the distance matrix (`"d"`) and the desired clustering function:

```{r crisp-silhouette5, fig.width=7, fig.height=4, fig.alt = "fig1.3"}
sil_fcm <- Silhouette(prox_matrix = "d", clust_fun = fcm, x = iris[, -5], centers = 3)
plot(sil_fcm)
```


This approach eliminates the explicit step of extracting the proximity matrix, making analyses more concise.

**Summary:**\
When the proximity matrix is provided directly by a clustering algorithm (as with fuzzy c-means), silhouette widths can be calculated in one step.
For further convenience, the `Silhouette()` function accepts both the proximity matrix and a clustering function, so that a single command completes the clustering and computes silhouettes.
This greatly simplifies the process for methods with built-in proximity outputs, supporting rapid and reproducible evaluation of clustering separation and quality.

### c. Calculation of Fuzzy Silhouette Index for Soft Clustering Algorithms

This section explains how to compute the fuzzy silhouette index when both the **proximity matrix** (distances from observations to cluster centers) and the **membership probability matrix** are available.
The process is demonstrated with fuzzy c-means clustering from the `ppclust` package applied to the classic `iris` dataset.

**Steps:**

- **Step 1: Perform Fuzzy C-Means Clustering**\
    Apply fuzzy c-means clustering to the feature columns of the `iris` dataset, specifying three clusters:

```{r fuzzy-silhouette4.1, fig.width=7, fig.height=4}
data(iris)
fm1 <- ppclust::fcm(x = iris[, -5], centers = 3)
```

- **Step 2: Compute Fuzzy Silhouette Widths Using Proximity and Membership Matrices**\
    The clustering output `fm1` contains both the distance matrix (`fm1$d`) and the membership probability matrix (`fm1$u`) and `average = "fuzzy"`. These can be directly passed to the `Silhouette()` function to compute fuzzy silhouette widths:

```{r fuzzy-silhouette4, fig.width=7, fig.height=4, fig.alt = "fig1.6"}
sil_fm1 <- Silhouette(prox_matrix = fm1$d, prob_matrix = fm1$u, average = "fuzzy")
plot(sil_fm1)
```


- **Alternative: Use Clustering Function Inline with `clust_fun`**\
    For an even more streamlined workflow, the `Silhouette()` function can internally manage clustering and silhouette calculations by accepting the names of the distance and probability components (`"d"` and `"u"`) along with the clustering function:

```{r fuzzy-silhouette5, fig.width=7, fig.height=4, fig.alt = "fig1.3"}
library(ppclust)
sil_fcm1 <- Silhouette(prox_matrix = "d", prob_matrix = "u", average = "fuzzy", clust_fun = fcm, x = iris[, -5], centers = 3)
plot(sil_fcm1)
```

This approach removes the need to manually extract matrices from the clustering result, improving code efficiency and reproducibility.

**Summary:**\
When both the proximity and membership probability matrices are directly available from a clustering algorithm (such as fuzzy c-means), fuzzy silhouette widths can be calculated efficiently in a single step.
The `Silhouette()` function further supports an integrated workflow by running both the clustering and silhouette calculations internally when provided with the relevant function and argument names.
This functionality facilitates a concise, reproducible pipeline for validating the quality and separation of soft clustering results.

## 2. Comparing Two Soft Clustering Algorithms Using the Soft Silhouette Function

It is often desirable to assess and compare the clustering quality of different soft clustering algorithms on the same dataset.
The **soft silhouette index** offers a principled, internal measure for this purpose, as it naturally incorporates the probabilistic nature of soft clusters and provides a single value summarizing both cluster compactness and separation.

**Example: Evaluating Fuzzy C-Means vs. an Alternative Soft Clustering Algorithm**

Suppose we wish to compare the performance of two fuzzy clustering algorithms—such as Fuzzy C-Means (FCM) and a variant (e.g., FCM2)—using the `softSilhouette()` function.

**Steps:**

- **Step 1: Perform Clustering with Both Algorithms**

    Fit each soft clustering algorithm on your dataset (e.g., `iris[, 1:4]`):

```{r fcm}
data(iris)

# FCM clustering
fcm_result <- ppclust::fcm(iris[, 1:4], 3)

# FCM2 clustering
fcm2_result <- ppclust::fcm2(iris[, 1:4], 3)
```

-   **Step 2: Compute Soft Silhouette Index for Each Result**

    Use the membership probability matrices produced by each algorithm:

```{r softSilhouette, fig.width=7, fig.height=4, fig.alt = "fig2.1"}
# Soft silhouette for FCM
sil_fcm <- softSilhouette(prob_matrix = fcm_result$u)
plot(sil_fcm)

# Soft silhouette for FCM2
sil_fcm2 <- softSilhouette(prob_matrix = fcm2_result$u)
plot(sil_fcm2)
```

-   **Step 3: Summarize and Compare Average Silhouette Widths**

    Extract the overall average silhouette width for each clustering result:

```{r softSilhouette1, fig.width=7, fig.height=4, fig.alt = "fig2.2"}
sfcm <- summary(sil_fcm, print.summary = FALSE)
sfcm2 <- summary(sil_fcm2, print.summary = FALSE)

cat("FCM average silhouette width:", sfcm$avg.width, "\n",
    "FCM2 average silhouette width:", sfcm2$avg.width)

```

A higher average silhouette width indicates a clustering with more compact and well-separated clusters.

**Interpretation & Guidance**

-   **Interpret the Index**: The algorithm yielding a higher *average soft silhouette width* is considered to produce a better clustering, as it balances cluster cohesion and separation while accounting for the uncertainty inherent in soft assignments.
-   **Practical Application**: This method is generic; any two or more soft clustering results (not limited to FCM/FCM2) can be compared effectively, provided you can extract the membership probability matrix.
-   **Flexible Integration**: The `softSilhouette()` function also allows for different silhouette calculation methods and transformations (such as `prob_type = "nlpp"` for negative log-probabilities), supporting deeper comparisons aligned with your methodological framework.

**Additional Soft Clustering Methods**

The package also provides two additional methods for computing soft silhouette widths: `cerSilhouette()` (Certainty-based) and `dbSilhouette()` (Density-based). These can be used in the same way as `softSilhouette()` to compare clustering algorithms.

```{r cer-db-silhouette, fig.width=7, fig.height=4, fig.alt = "fig2.3"}
# Certainty-based silhouette for FCM and FCM2
  cer_fcm <- cerSilhouette(prob_matrix = fcm_result$u, print.summary = TRUE)
  plot(cer_fcm)
  
  cer_fcm2 <- cerSilhouette(prob_matrix = fcm2_result$u, print.summary = TRUE)
  plot(cer_fcm2)

# Density-based silhouette for FCM and FCM2
  db_fcm <- dbSilhouette(prob_matrix = fcm_result$u, print.summary = TRUE)
  plot(db_fcm)
  
  db_fcm2 <- dbSilhouette(prob_matrix = fcm2_result$u, print.summary = TRUE)
  plot(db_fcm2)
```

```{r cer-db-silhouette-summary}
# Compare average silhouette widths across all methods
  # Summary for FCM
  cer_sfcm <- summary(cer_fcm, print.summary = FALSE)
  db_sfcm <- summary(db_fcm, print.summary = FALSE)
  
  # Summary for FCM2
  cer_sfcm2 <- summary(cer_fcm2, print.summary = FALSE)
  db_sfcm2 <- summary(db_fcm2, print.summary = FALSE)
  
  # Print comparison
  cat("FCM - Soft silhouette:", sfcm$avg.width, "\n",
      "FCM - Certainty silhouette:", cer_sfcm$avg.width, "\n",
      "FCM - Density-based silhouette:", db_sfcm$avg.width, 
      "\n\n","FCM2 - Soft silhouette:", sfcm2$avg.width, 
      "\n","FCM2 - Certainty silhouette:", cer_sfcm2$avg.width, 
      "\n","FCM2 - Density-based silhouette:", db_sfcm2$avg.width, "\n")
```

**Summary:**\
Comparing the average soft silhouette widths from different soft clustering algorithms provides an objective, data-driven basis for determining which method produces more meaningful, well-defined clusters in probabilistic settings.
This approach harmonizes easily with both classic fuzzy clustering and more advanced algorithms, and can be extended to other soft silhouette methods like certainty-based and density-based approaches.

## 3. Scree Plot for Optimal Number of Clusters

The scree plot (also called the "elbow plot" or "reverse elbow plot") is a practical tool for identifying the best number of clusters in unsupervised learning.
Here, the silhouette width is calculated for different values of *k* (number of clusters).
The resulting plot provides a visual indication of the optimal cluster count by highlighting where increasing *k* yields only marginal improvements in the average silhouette width.

**Steps:**

-   **Step 1: Compute Average Silhouette Widths at Varying Cluster Counts**\
    Run silhouette analysis across a range of possible cluster numbers (e.g., 2 to 7). For each *k*, use the any `Silhouette` class function to calculate the silhouette widths, then extract the average silhouette width from the summary.

```{r screeplot1}
data(iris)
avg_sil_width <- rep(NA,7)
for (k in 2:7) {
  sil_out <- Silhouette(
    prox_matrix = "d",
    method = "pac",
    clust_fun = ppclust::fcm,
    x = iris[, 1:4],
    centers = k)
  avg_sil_width[k] <- summary(sil_out, print.summary = FALSE)$avg.width
}
```

-   **Step 2: Create and Interpret the Scree Plot**\
    Plot the number of clusters against the computed average silhouette widths:

```{r screeplot2, fig.width=7, fig.height=4, fig.alt = "fig3.1"}
plot(avg_sil_width,
  type = "o",
  ylab = "Overall Silhouette Width",
  xlab = "Number of Clusters",
  main = "Silhouette Scree Plot"
)
```

The optimal number of clusters is often suggested by the "elbow" or "reverse elbow"—the point after which increases in *k* lead to diminishing or excessive improvements in silhouette width.
This visual guide is valuable for assessing the clustering structure in your data.

*Note:* Any `Silhouette` class functions can be used to generate scree plots for optimal cluster selection.
For theoretical background and additional diagnostic options for soft clustering, see @bhatkapu2024density.

**Summary:**\
The scree plot provides an intuitive graphical summary to assist in choosing the optimal number of clusters by plotting average silhouette width versus the number of clusters considered.
The integrated use of `Silhouette()`, `softSilhouette()`, `cerSilhouette()`, `dbSilhouette()` use of `clust_fun` and summary functions makes this analysis straightforward and efficient for both crisp and fuzzy clustering frameworks.
This method encourages a reproducible, objective approach to cluster selection in unsupervised analysis.

## 4. Visualizing Silhouette Analysis Results with `plotSilhouette()`

Efficient visualization of silhouette widths is essential for interpreting and diagnosing clustering quality. The `plotSilhouette()` function provides a flexible and extensible tool for plotting silhouette results from various clustering algorithms, supporting both hard (crisp) and soft (fuzzy) partitions.

**Key Features:**
- Accepts outputs from a wide range of clustering methods: `Silhouette`, `softSilhouette`, `dbSilhouette`, `cerSilhouette` as well as clustering objects from `cluster` (`pam`, `clara`, `fanny`, base `silhouette`) and `factoextra` (`eclust`, `hcut`).
- Offers detailed legends summarizing average silhouette widths and cluster sizes.
- Supports customizable color palettes, including grayscale, and the option to label observations on the x-axis.

**Illustrative Use Cases and Code**

- **Crisp Silhouette Visualization (e.g., k-means clustering):**

```{r plot0, fig.width=6, fig.height=4, fig.alt = "fig4.0"}
data(iris)
  km_out <- kmeans(iris[, -5], 3)
  dist_mat <- proxy::dist(iris[, -5], km_out$centers)
  sil_obj <- Silhouette(dist_mat)
  plot(sil_obj)                   # S3 method auto-dispatch
  plotSilhouette(sil_obj)         # explicit call (identical output)
```

- **Crisp Silhouette from Cluster Algorithms (PAM, CLARA, FANNY):**

```{r plot1, fig.width=6, fig.height=4, fig.alt = "fig4.1"}
library(cluster)
pam_result <- pam(iris[, 1:4], k = 3)
plotSilhouette(pam_result) # for cluster::pam object

clara_result <- clara(iris[, 1:4], k = 3)
plotSilhouette(clara_result)

fanny_result <- fanny(iris[, 1:4], k = 3)
plotSilhouette(fanny_result)
```

- **Base silhouette object:**

```{r plot2, fig.width=6, fig.height=4, fig.alt = "fig4.2"}  
sil_base <- cluster::silhouette(pam_result)
plotSilhouette(sil_base)
```

- **factoextra::hcut/eclust clusterings:**

```{r plot3, fig.width=6, fig.height=4, fig.alt = "fig4.3"}
library(factoextra)
eclust_result <- eclust(iris[, 1:4], "kmeans", k = 3, graph = FALSE)
plotSilhouette(eclust_result)

hcut_result <- hcut(iris[, 1:4], k = 3)
plotSilhouette(hcut_result)
```

- **drclust::silhouette Visualization:**

```{r plot3.1, fig.width=6, fig.height=5, fig.alt = "fig4.3.1"}
library(drclust)
# Loading the numeric in matrix 
iris_mat <- as.matrix(iris[,-5])
#applying a clustering algorithm
drclust_out <- dpcakm(iris_mat, 20, 3)
#silhouette based on the data and the output of the clustering algorithm
d <- silhouette(iris_mat, drclust_out)
plotSilhouette(d$cl.silhouette)
```

- **Fuzzy (Soft) Silhouette Visualization (e.g., fuzzy c-means with ppclust):**

```{r plot4, fig.width=6, fig.height=4, fig.alt = "fig4.4"}
data(iris)
fcm_out <- ppclust::fcm(iris[, 1:4], 3)
sil_fuzzy <- Silhouette(
  prox_matrix = "d", prob_matrix = "u", clust_fun = fcm,
  x = iris[, 1:4], centers = 3, sort = TRUE
)
plot(sil_fuzzy, summary.legend = FALSE, grayscale = TRUE)
```

- **Customization: Grayscale, Detailed Legends, and Observation Labels:**

```{r plot5, fig.width=6, fig.height=4, fig.alt = "fig4.5"}
plotSilhouette(sil_fuzzy, grayscale = TRUE) # Use grayscale palette
plotSilhouette(sil_fuzzy, summary.legend = TRUE) # Include size + avg silhouette in legend
plotSilhouette(sil_fuzzy, label = TRUE) # Label bars with row index
```

**Practical Guidance:**
- For clustering output classes not supported by the generic `plot()` function, always use `plotSilhouette()` explicitly to ensure correct and informative visualization.
- The function automatically sorts silhouette widths within clusters, displays the average silhouette (dashed line), and provides detailed cluster summaries in the legend.

**Summary:**  
`plotSilhouette()` brings unified, publication-ready visualization capabilities for assessing crisp and fuzzy clustering at a glance. Its broad compatibility, detailed legends, grayscale and labeling options empower users to gain deeper insights into clustering structure, facilitating clear diagnosis and reporting in both exploratory and formal statistical workflows.

## 5. Creating and Validating User-Defined Silhouette Objects

The `getSilhouette()` function enables users to manually construct a `Silhouette` object from precomputed cluster assignments, neighbor clusters, silhouette widths, and optional weights.
This is particularly useful for custom or externally derived clustering results.

The `is.Silhouette()` function validates whether an object is a valid `Silhouette` object, ensuring it meets the necessary structural and attribute requirements for visualization and analysis.

Example:
```{r custom1, fig.width=7, fig.height=4, fig.alt = "fig5.1"}
# Create a custom Silhouette object
cluster_assignments <- c(1, 1, 2, 2, 3, 3)
neighbor_clusters <- c(2, 2, 1, 1, 1, 1)
silhouette_widths <- c(0.8, 0.7, 0.6, 0.9, 0.5, 0.4)
weights <- c(0.9, 0.8, 0.7, 0.95, 0.6, 0.5)

sil_custom <- getSilhouette(
  cluster = cluster_assignments,
  neighbor = neighbor_clusters,
  sil_width = silhouette_widths,
  weight = weights,
  proximity_type = "similarity",
  method = "pac",
  average = "fuzzy"
)
# Validate the object
is.Silhouette(sil_custom)               # Basic class check: TRUE
is.Silhouette(sil_custom, strict = TRUE) # Strict structural validation: TRUE
is.Silhouette(data.frame(a = 1:6))      # Non-Silhouette object: FALSE
# Visualize the custom Silhouette object
plotSilhouette(sil_custom, summary.legend = TRUE)
```

This approach allows users to integrate custom silhouette computations into the **Silhouette** package’s visualization framework, ensuring flexibility for specialized workflows while maintaining compatibility with `plotSilhouette()`.

## 6. Extended Silhouette Analysis for Multi-Way Clustering

The `extSilhouette()` function enables silhouette-based evaluation for multi-way clustering scenarios, such as biclustering or tensor clustering, by aggregating silhouette indices from each mode (e.g., rows, columns) into a single summary metric.
This approach allows you to rigorously assess the overall clustering structure when partitioning data along multiple dimensions.

**Workflow:**

-   **Step 1: Apply Multi-Way Clustering**\
    Fit a biclustering algorithm to your data—in this example, we use `blockcluster::coclusterContinuous()` to jointly cluster the rows and columns of the `iris` dataset.

```{r ext1}
library(blockcluster)
data(iris)
result <- coclusterContinuous(as.matrix(iris[, -5]), nbcocluster = c(3, 2))
```

-   **Step 2: Compute Silhouette Widths for Each Mode**\
    For each dimension (e.g., rows and columns), calculate silhouette widths using the membership probability matrices (`result@rowposteriorprob` for rows, `result@colposteriorprob` for columns) via the `softSilhouette()` function: (One can use any `Silhouette` class function to calculate when relevant proximity measure available, For consistency make sure all \code{Silhouette} objects in list derived from same `method` and arguments.)

```{r ext2}
sil_mode1 <- softSilhouette(
  prob_matrix = result@rowposteriorprob,
  method = "pac",
  print.summary = FALSE
)
sil_mode2 <- softSilhouette(
  prob_matrix = result@colposteriorprob,
  method = "pac",
  print.summary = FALSE
)
```

-   **Step 3: Aggregate Silhouette Results with `extSilhouette()`**\
    Combine the silhouette analyses from each mode by passing them as a list to `extSilhouette()`. Optionally, provide descriptive dimension names:

```{r ext3}
ext_sil <- extSilhouette(
  sil_list = list(sil_mode1, sil_mode2),
  dim_names = c("Rows", "Columns"),
  print.summary = TRUE
)
```

**Summary:**\
The `extSilhouette()` function returns: - The overall extended silhouette width—a weighted average summarizing clustering quality across all modes.
- A dimension statistics table, reporting the number of observations and average silhouette width for each mode (e.g., rows, columns).

*Note:*\
If a distance matrix is available from the output of a biclustering algorithm, you can compute individual mode silhouettes using `Silhouette()`.

The results can be combined with `extSilhouette()` to enable direct comparison of clustering solutions across multiple biclustering algorithms, facilitating objective model assessment [@bhat2025block].

This methodology provides a concise and interpretable assessment for complex clustering models where conventional one-dimensional indices are insufficient.

# References
